
# Working with Factors/Categorical Data {#factors}

<script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script>

```{r include=FALSE}
tutorial::go_interactive(greedy = FALSE)
```

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this section, we'll learn some skills about manipulating factor (categorical) data.

We'll do this by making a bar plot and a box plot and progressively making it more complex.

## Reviewing Factors 

Factors are how R represents categorical data.

There are two kinds of factors: 

+ `factor` - used for *nominal* data ("Ducks","Cats","Dogs")
+ `ordered`- used for *ordinal* data ("10-30","31-40","41-60")

We'll manipulate our barplot and add more information using factors.

Here's the simple dataset we'll use to investigate how to work with factors in `ggplot`.

For the `factor` and `ordered` variables, find the categories for each using `levels()`.

```{r}
library(tidyverse)
load("data/pets.rda")
pets
```

## A Basic Barplot using `geom_bar()`

The `geom_bar()` default is to count the number of values with each factor level. Note that you don't map to a y-aesthetic here, because the y values are the counts.

Given this dataset, we might want to ask how many pets have the same name. What is the most popular name?

Try mapping another variable to `fill` (try both `weight` and `animal`). Are the results what you expected?

```{r}
ggplot(data = pets, mapping = aes(x=name)) + geom_bar()
```

## Faceting

Say you have another `factor` variable and you want to stratify the plots based on that. 
You can do that by supplying the name of that variable as a facet. Here, we facet our barplot by `shotsCurrent`.

```{r}
ggplot(data = pets, mapping = aes(x = name)) + geom_bar() + 
  ## have to specify facets using notation
  ## try out facets=~ageCategory!
  facet_wrap(facets = ~shotsCurrent) + 
  ## we make the x axis x angled for better legibility
  theme(axis.text.x = element_text(angle = 45))
```

You might notice that there are blank spots for the categories in each facet. We can restrict the factors in each by using `scale="free_x"` argument in `facet_wrap()`.

How many animals named "Morris" did not receive shots?

What happens when you replace the `scale` argument with "free_y"?

```{r}
ggplot(pets, aes(x = name)) + geom_bar() + 
  facet_wrap(facets = ~shotsCurrent, scale = "free_x") +
  theme(axis.text.x = element_text(angle = 45))
```

## Stacked Bars

Let's see how many of each animal got shots. We can do this by mapping `shotsCurrent` to `fill`.

```{r}
#we map color (the outline of the plot) to black to make it look prettier
ggplot(pets, aes(x = animal, fill = shotsCurrent)) + 
  geom_bar(color = "black")
```

## Proportional Barchart

We may only be interested in the relative proportions between the different categories. Visualizing this is useful for various 2 x 2 tests on proportions.

What percent of dogs did not receive shots?

```{r}
ggplot(pets, aes(x = animal, fill = shotsCurrent)) +
  geom_bar(position = "fill", color = "black")
```

## Dodge those bars!

Instead of stacking, we can also dodge the bars (move the bars so they're beside each other).

```{r}
ggplot(pets, aes(x = animal, fill = shotsCurrent)) +
  geom_bar(position = "dodge", color = "black")
```

## Your Task: Bar Charts

Given the `pets` `data.frame`, plot a stacked proportional barchart that shows the age category counts by animal type. Is the proportion of animals receiving shots the same across each age category?

Hints: think about what to map to `x`, and what to map to `fill`.

```{r ex="pets_bar", type="pre-exercise-code", tut=TRUE}
pets <- data.frame(id = c('C1','C2','D1','D4','D8', 'C4', 'D5', 
                        'C10', "G4", "C11", "D9", "C6"),
                   name = factor(c("Morris", "Morris", "Fido", "Mr Bowser", 
                                 "Fido", "Lady Sheba", "Hubert", "Apples", "Nibbles",
                                 "Morris", "Winky", "Fido")),
                   ##weight is in pounds
                   weight = c(6, 7, 15, 20, 40, 10, 60, 17, 0.2, 10, 11, 6),
                   animal = factor(c("cat", "cat", "dog", "dog", "dog", 
                                   "cat", "dog", "cat", "gerbil", "cat", "dog", "cat")),
                   ##because it's an ordered category, we have to specify 
                   ##order as the levels argument
                   ageCategory = ordered(c("0-2", "5-8", "5-8", "3-4", "3-4", 
                                          "5-8", "0-2", "5-8", "0-2","5-8", "3-4", "5-8"), 
                                        levels = c("0-2", "3-4", "5-8")),
                   shotsCurrent = c("No", "Yes", "No", "No", "Yes", "No", 
                                  "No", "Yes", NA, "No", "No", "Yes"))
```

```{r ex="pets_bar", type="sample-code", tut=TRUE}
library(ggplot2)

#Space for your answer here.

```

```{r ex="pets_bar", type="solution", tut=TRUE}
library(ggplot2)
ggplot(data = pets, 
       mapping = aes(x = ageCategory, fill = animal)) +
  geom_bar(position = "fill")
```

```{r ex="pets_bar", type="sct", tut=TRUE}
test_library_function("ggplot2")
test_ggplot(check_data = TRUE, check_aes = TRUE, check_geom = TRUE)
success_msg("Nice!")
test_error()
```

**Intermediate Folks: facet this plot by ** `shotsCurrent` **using** `facet_wrap`. 


```{r ex="pets_bar_int", type="pre-exercise-code", tut=TRUE}
pets <- data.frame(id = c('C1','C2','D1','D4','D8', 'C4', 'D5', 
                        'C10', "G4", "C11", "D9", "C6"),
                   name = factor(c("Morris", "Morris", "Fido", "Mr Bowser", 
                                 "Fido", "Lady Sheba", "Hubert", "Apples", "Nibbles",
                                 "Morris", "Winky", "Fido")),
                   ##weight is in pounds
                   weight = c(6, 7, 15, 20, 40, 10, 60, 17, 0.2, 10, 11, 6),
                   animal = factor(c("cat", "cat", "dog", "dog", "dog", 
                                   "cat", "dog", "cat", "gerbil", "cat", "dog", "cat")),
                   ##because it's an ordered category, we have to specify 
                   ##order as the levels argument
                   ageCategory = ordered(c("0-2", "5-8", "5-8", "3-4", "3-4", 
                                          "5-8", "0-2", "5-8", "0-2","5-8", "3-4", "5-8"), 
                                        levels = c("0-2", "3-4", "5-8")),
                   shotsCurrent = c("No", "Yes", "No", "No", "Yes", "No", 
                                  "No", "Yes", NA, "No", "No", "Yes"))
```

```{r ex="pets_bar_int", type="sample-code", tut=TRUE}
library(ggplot2)

#Space for your answer here.

```

```{r ex="pets_bar_int", type="solution", tut=TRUE}
library(ggplot2)
ggplot(data = pets, 
       mapping = aes(x = ageCategory, fill = animal)) +
  geom_bar(position = "fill") +
  facet_wrap(facets = ~shotsCurrent)
```

```{r ex="pets_bar_int", type="sct", tut=TRUE}
test_library_function("ggplot2")
test_ggplot(check_data = TRUE, check_facet = TRUE, check_aes = TRUE, 
            facet_fail_msg = "You didn't map the correct facet. Try again.", 
            aes_fail_msg = "Not the correct fill aesthetic. Try again.")
success_msg("Nice!")
test_error()
```

## Boxplots

Boxplots allow us to assess distributions of a continuous variable conditioned on categorical variables.

What does this tell us? 

```{r}
ggplot(pets, aes(x = shotsCurrent, y = weight)) + 
  geom_boxplot()
```

## Violin Plots

Violin plots are another useful way to visualize the data. They provide a more nuanced look at the data. They're a density plot that's mirrored around the y-axis.

```{r}
ggplot(pets, aes(x = ageCategory, y = weight, fill = ageCategory)) + 
  geom_violin()
```

## Your task: How heavy are our pets?

Visualize `weight` by `animal` type as both a boxplot and a violin plot on the same plot. What do you conclude? Which kind of animal weighs more on average than the other?

```{r ex="box_violin", type="pre-exercise-code", tut=TRUE}
pets <- data.frame(id = c('C1','C2','D1','D4','D8', 'C4', 'D5', 
                        'C10', "G4", "C11", "D9", "C6"),
                   name = factor(c("Morris", "Morris", "Fido", "Mr Bowser", 
                                 "Fido", "Lady Sheba", "Hubert", "Apples", "Nibbles",
                                 "Morris", "Winky", "Fido")),
                   ##weight is in pounds
                   weight = c(6, 7, 15, 20, 40, 10, 60, 17, 0.2, 10, 11, 6),
                   animal = factor(c("cat", "cat", "dog", "dog", "dog", 
                                   "cat", "dog", "cat", "gerbil", "cat", "dog", "cat")),
                   ##because it's an ordered category, we have to specify 
                   ##order as the levels argument
                   ageCategory = ordered(c("0-2", "5-8", "5-8", "3-4", "3-4", 
                                          "5-8", "0-2", "5-8", "0-2","5-8", "3-4", "5-8"), 
                                        levels = c("0-2", "3-4", "5-8")),
                   shotsCurrent = c("No", "Yes", "No", "No", "Yes", "No", 
                                  "No", "Yes", NA, "No", "No", "Yes"))

```

```{r ex="box_violin", type="sample-code", tut=TRUE}
library(ggplot2)

```

```{r ex="box_violin", type="solution", tut=TRUE}
library(ggplot2)
ggplot(data = pets, mapping = aes(x = animal, y = weight)) +
  geom_boxplot() + 
  geom_violin()
```

```{r ex="box_violin", type="sct", tut=TRUE}
test_library_function("ggplot2")
test_ggplot(check_aes = TRUE, aes_fail_msg = "Not quite. Check your mapping")
test_function("geom_boxplot", incorrect_msg = "I think you're missing a geom here.")
test_function("geom_violin", incorrect_msg = "I think you're missing a geom.")
success_msg("Great! You can see that the boxplot and the violin plot contribute different levels of detail to the graph")
```


## What you learned in this section

- Visualizing factor data
- Simple, stacked, stacked proportional, and dodged barplots 
- Faceting a graph
- Boxplots and violin plots
